---
title: "Wastewater informed forecasting: fit to simulated data"
author: "Kaitlyn Johnson"
date: "2023-12-20"
output: html_document
---

This vignette steps through the pre-processing and fitting procedure to forecast
state-level hospital admissions using simulated data. The code can be adapted
for your own datasets by replacing the joined hospital admissions + site-lab
level wastewater dataset, here called, `example_df`, with your own dataset and
varying the parameters that are pre-specified. See the documentation for
`example_df` to format your data in a similar manner. For an example of how to
do this from the raw generated outputs, see the `generate_simulated_data.R`
function. Note that if using real data, the part about comparing the inferred
parameters to the known truth parameters will not be relevant as this information will not be available.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(cfaforecastrenewalww)
library(here)
library(dplyr)
library(ggplot2)
library(tidybayes)
```
# Rationale


# Load the data
```{r}
# Example of how to generate new simulated data with user-specified parameters
# by calling `generate_simulated_data()`
# See function documentation to modify R(t), number of sites, etc.
example_df <- cfaforecastrenewalww::example_df
```
The `generate_simulated_data.R` file can be used to generate this simulated
data where individual sites exhibit their own site-level infection dynamics.
Setting `n_sites` to one
allows the user to generate data amenable to the state-level aggregated
wastewater model, or alternatively, site-level data can be generated and then
aggregated using population weighted averaging of the concentration to generate
a single state-level wastewater data stream. However, this vignette assumes you
have wastewater data from one or more sites that represent some subset of the
population for which you have hospital admissions data for.

# Data exploration
The example dataframe is a long tidy data frame with hospital admissions alongside
site-level wastewater data, corresponding to a single observation for each
lab-site-day. We assign an arbitrary date to the time series so that we can
assign a week day to each day. We additionally include a column containing
the daily hospital admissions through the forecast period
`daily_hosp_admits_for_eval`. This will
be used for evaluating our simulated data inference procedure.
See documentation of `example_df` for a full description of each column.
```{r}
head(example_df)
```
# Make some plots of the hospital admissions and wastewater data
The model will
jointly calibrate to state-level hospital admissions and the site-level
wastewater data. It will estimate underlying latent incident infection curves,
either for each site coming from a shared distribution with a global mean of the
state-level incident infection curve, or it will estimate a single state-level
incident infection curve that is assumes to be homogenous across all sites.
From the simulated data, we impose a delay from the date of the forecast to the
last observed hospital admissions, to reflect the current data reporting
structure in the US. We also include variable reporting frequency and latency
in the site-level wastewater data, to reflect the real heterogeneity we observe
in the NWSS data in real-time.
```{r}
ggplot(example_df) +
  geom_point(
    aes(
      x = date, y = exp(log_conc),
      color = as.factor(lab_wwtp_unique_id)
    ),
    show.legend = FALSE
  ) +
  geom_point(
    data = example_df %>% filter(below_LOD == 1),
    aes(x = date, y = exp(log_conc), color = "red"),
    show.legend = FALSE
  ) +
  geom_hline(aes(yintercept = exp(lod_sewage)), linetype = "dashed") +
  facet_wrap(~lab_wwtp_unique_id, scales = "free") +
  geom_vline(aes(xintercept = forecast_date), linetype = "dashed") +
  xlab("") +
  ylab("Genome copies/mL") +
  ggtitle("Lab-site level wastewater concentration") +
  theme_bw()

ggplot(example_df) +
  geom_point(aes(x = date, y = daily_hosp_admits_for_eval),
    shape = 21, color = "black", fill = "white"
  ) +
  geom_point(aes(x = date, y = daily_hosp_admits)) +
  geom_vline(aes(xintercept = forecast_date), linetype = "dashed") +
  xlab("") +
  ylab("Daily hospital admissions") +
  ggtitle("State level hospital admissions") +
  theme_bw()
```

# Format data for inference in stan
The model expects input data in a certain format. The `cfaforecastrenewalww`
R package provides helper functions to facilitate preparing data for ingestion by the model.

Hyperparameters for prior distributions and other fixed quantites should be specified in a `.toml` file, which the function `get_params()` reads and validates. We have provided an example parameter file, `example_params.toml`, with the package.

We'll read it in and examine it here:

```{r}
params <- get_params(
  system.file("extdata", "example_params.toml",
    package = "cfaforecastrenewalww"
  )
)
params
```

The example values in `example_params.toml` are the same as those stored in `cfaforecastrenewalww::param_df` and used (via prior predictive simulation to generate the simulated data in `cfaforecastrenewalww::example_df`.

```{r}
forecast_date <- example_df %>%
  pull(forecast_date) %>%
  unique()
forecast_time <- as.integer(max(example_df$date) - forecast_date)

# Assign the model that we want to fit to  so that we grab the correct
# model and initialization list
model_type <- "site-level infection dynamics"
model_file_path <- get_model_file_path(model_type)
# Compile the model
model <- compile_model(file.path(here(
  model_file_path
)))


# Function calls for linear scale ww data
train_data_raw <- example_df %>%
  mutate(
    ww = exp(log_conc),
    period = case_when(
      !is.na(daily_hosp_admits) ~ "calibration",
      is.na(daily_hosp_admits) & date <= forecast_date ~
        "nowcast",
      TRUE ~ "forecast"
    ),
    include_ww = 1,
    site_index = site,
    lab_site_index = lab_wwtp_unique_id
  )

# Apply outliers to data
train_data <- flag_ww_outliers(train_data_raw)

# Get the generation interval and time from infection to hospital admission
# delay distribution to pass to stan.
# Use the same values as we do in the
# generation of these vectors in the generate simulated data function....
# See Park et al
# https://www.medrxiv.org/content/10.1101/2024.01.12.24301247v1 for why
# we use a double-censored pmf here
generation_interval <- simulate_double_censored_pmf(
  max = params$gt_max, meanlog = params$mu_gi,
  sdlog = params$sigma_gi, fun_dist = rlnorm, n = 5e6
) %>% drop_first_and_renormalize()

inc <- make_incubation_period_pmf(
  params$backward_scale, params$backward_shape, params$r
)
sym_to_hosp <- make_hospital_onset_delay_pmf(
  params$neg_binom_mu,
  params$neg_binom_size
)
inf_to_hosp <- make_reporting_delay_pmf(inc, sym_to_hosp)

# Format as a list for stan
stan_data <- get_stan_data_site_level_model(
  train_data,
  params,
  forecast_date,
  forecast_time,
  model_type = model_type,
  generation_interval = generation_interval,
  inf_to_hosp = inf_to_hosp,
  infection_feedback_pmf = generation_interval
)

init_fun <- function() {
  site_level_inf_inits(train_data, params, stan_data)
}
```
# Fit the wastewaster informed model
Here we will use MCMC settings with 500 warmup iterations and 750 sampling
iterations, with settings for the maximum treedepth (referred to as
`max_treedepth`) of 12 and an acceptance probability (referred to as
`adapt_delta`) of 0.95. This matches the production-level settings specified in
`src/write_config.R`.
```{r warning=FALSE, message=FALSE}
fit_dynamic_rt <- model$sample(
  data = stan_data,
  seed = 123,
  init = init_fun,
  iter_sampling = 500,
  iter_warmup = 750,
  max_treedepth = 12,
  chains = 4,
  parallel_chains = 4
)
```

# Look at the model outputs: generated quantities
Plot the model predicted hospital admissions vs the simulated forecasted
hospital admissions and the model estimated wastewater concentration in each
site vs the simulated wastewater data.
```{r}
all_draws <- fit_dynamic_rt$draws()


# Predicted observed hospital admissions
exp_obs_hosp <- all_draws %>%
  spread_draws(pred_hosp[t]) %>%
  select(pred_hosp, `.draw`, t) %>%
  rename(draw = `.draw`)
sampled_draws <- sample(1:max(exp_obs_hosp$draw), 100)

# Predicted observed wastewater concentrations in each lab site
exp_obs_conc <- all_draws %>%
  spread_draws(pred_ww[lab_wwtp_unique_id, t]) %>%
  select(pred_ww, `.draw`, t, lab_wwtp_unique_id) %>%
  rename(draw = `.draw`)

output_df <- exp_obs_conc %>%
  left_join(train_data,
    by = c("t", "lab_wwtp_unique_id")
  ) %>%
  left_join(exp_obs_hosp,
    by = c("t", "draw")
  ) %>%
  filter(draw %in% sampled_draws) # sample the draws for plotting

ggplot(output_df) +
  geom_line(aes(x = date, y = pred_hosp, group = draw),
    color = "red4", alpha = 0.1, size = 0.2
  ) +
  geom_point(aes(x = date, y = daily_hosp_admits_for_eval),
    shape = 21, color = "black", fill = "white"
  ) +
  geom_point(aes(x = date, y = daily_hosp_admits)) +
  geom_vline(aes(xintercept = forecast_date), linetype = "dashed") +
  xlab("") +
  ylab("Daily hospital admissions") +
  ggtitle("State level hospital admissions estimated with wastewater") +
  theme_bw()

ggplot(output_df) +
  geom_line(
    aes(
      x = date, y = exp(pred_ww),
      color = as.factor(lab_wwtp_unique_id),
      group = draw
    ),
    alpha = 0.1, size = 0.2,
    show.legend = FALSE
  ) +
  geom_point(aes(x = date, y = exp(log_conc)),
    color = "black", show.legend = FALSE
  ) +
  geom_point(
    data = example_df %>% filter(below_LOD == 1),
    aes(x = date, y = exp(log_conc), color = "red"),
    show.legend = FALSE
  ) +
  geom_hline(aes(yintercept = exp(lod_sewage)), linetype = "dashed") +
  facet_wrap(~lab_wwtp_unique_id, scales = "free") +
  geom_vline(aes(xintercept = forecast_date), linetype = "dashed") +
  xlab("") +
  ylab("Genome copies/mL") +
  ggtitle("Lab-site level wastewater concentration") +
  theme_bw()

ggplot(output_df) +
  geom_line(
    aes(
      x = date, y = pred_ww,
      color = as.factor(lab_wwtp_unique_id),
      group = draw
    ),
    alpha = 0.1, size = 0.2,
    show.legend = FALSE
  ) +
  geom_point(aes(x = date, y = log_conc),
    color = "black", show.legend = FALSE
  ) +
  geom_point(
    data = example_df %>% filter(below_LOD == 1),
    aes(x = date, y = log_conc, color = "red"),
    show.legend = FALSE
  ) +
  geom_hline(aes(yintercept = lod_sewage), linetype = "dashed") +
  facet_wrap(~lab_wwtp_unique_id, scales = "free") +
  geom_vline(aes(xintercept = forecast_date), linetype = "dashed") +
  xlab("") +
  ylab("Log(genome copies/mL)") +
  ggtitle("Lab-site level wastewater concentration") +
  theme_bw()
```
<br>

# Model outputs: compare inferred parameters to known parameters
Get a dataframe of all of the key estimated model parameters combined with the key
"known" model parameters to get a sense of
which parameters we can recover and which might be unidentifiable (just as a
first pass, eventually we will probably want to do a full sensitivity analysis
to identify the regions of parameter space that are non-identifiable).

Note that if you are modifying this vignette to run on your own data skip this
part-- in the case of real data, we don't know the true underlying model
parameters.
```{r}
static_params <- c(
  "eta_sd", "autoreg_rt",
  "autoreg_rt_site",
  "log_r_mu_intercept",
  "i0_over_n", "initial_growth",
  "inv_sqrt_phi_h", "mode_sigma_ww_site", "sd_log_sigma_ww_site",
  "p_hosp_mean", "p_hosp_w_sd", "t_peak", "viral_peak",
  "dur_shed", "log10_g", "ww_site_mod_sd",
  "infection_feedback"
)
vector_params <- c(
  "w", "p_hosp_w", "rt", "state_inf_per_capita", "p_hosp",
  "eta_log_sigma_ww_site", "ww_site_mod_raw",
  "ww_site_mod", "sigma_ww_site",
  "hosp_wday_effect"
)

matrix_params <- c("error_site")


# Get the full posterior parameter distribution from the real data
full_param_df <- get_full_param_distrib(
  all_draws, static_params, vector_params, matrix_params
)

# Compare to the knows parameters
param_df <- cfaforecastrenewalww::param_df # A set of the set static and dynamic
# from the `generate_simulated_data()` function default args. This is package data
comp_df <- full_param_df %>%
  left_join(param_df,
    by = c("name", "index_rows", "index_cols")
  )

ggplot(comp_df %>% filter(name %in% c("rt", "p_hosp"))) +
  geom_line(aes(x = index_cols, y = value, group = draw), size = 0.05, alpha = 0.05) +
  geom_line(aes(x = index_cols, y = median), color = "black") +
  geom_line(aes(x = index_cols, y = true_value), color = "red") +
  facet_wrap(~name, scales = "free") +
  theme_bw() +
  xlab("Time (days)") +
  ylab("Parameter value") +
  scale_y_continuous(trans = "log") +
  ggtitle("Posterior parameter distribution vs known parameter values")

ggplot(comp_df %>% filter(name %in% c("log10_g"))) +
  geom_histogram(aes(x = value), alpha = 0.3) +
  geom_vline(aes(xintercept = median), color = "black") +
  geom_vline(aes(xintercept = true_value), color = "red") +
  facet_wrap(~name, scales = "free") +
  xlab("Parameter value") +
  ylab("Frequency") +
  theme_bw() +
  scale_x_continuous(trans = "log") +
  ggtitle("Posterior parameter distribution vs known parameter values")
```

# Try fitting the model without wastewater
The switch `include_ww = 0` tells the model not to evaluate the likelihood on
the wastewater data.
```{r warning=FALSE, message=FALSE}
stan_data_hosp_only <- stan_data
stan_data_hosp_only$include_ww <- 0

fit_dynamic_rt_hosp_only <- model$sample(
  data = stan_data_hosp_only,
  seed = 123,
  init = init_fun,
  iter_sampling = 500,
  iter_warmup = 750,
  chains = 4,
  max_treedepth = 12,
  parallel_chains = 4
)
```
```{r}
all_draws_hosp_only <- fit_dynamic_rt_hosp_only$draws()

# Predicted observed hospital admissions
exp_obs_hosp <- all_draws_hosp_only %>%
  spread_draws(pred_hosp[t]) %>%
  select(pred_hosp, `.draw`, t) %>%
  rename(draw = `.draw`)
sampled_draws <- sample(1:max(exp_obs_hosp$draw), 100)


output_df_hosp_only <- exp_obs_hosp %>%
  left_join(
    train_data %>%
      select(
        date, t, daily_hosp_admits,
        daily_hosp_admits_for_eval
      ) %>%
      distinct(),
    by = c("t")
  ) %>%
  filter(draw %in% sampled_draws) # sample the draws for plotting

ggplot(output_df_hosp_only) +
  geom_line(aes(x = date, y = pred_hosp, group = draw),
    color = "red4", alpha = 0.1, size = 0.2
  ) +
  geom_point(aes(x = date, y = daily_hosp_admits_for_eval),
    shape = 21, color = "black", fill = "white"
  ) +
  geom_point(aes(x = date, y = daily_hosp_admits)) +
  geom_vline(aes(xintercept = forecast_date), linetype = "dashed") +
  xlab("") +
  ylab("Daily hospital admssions") +
  ggtitle("State level hospital admissions fit to only hospital admissions") +
  theme_bw()
```

# Model outputs: compare inferred parameters to known parameters

This time just for the IHR and R(t)

```{r}
# Get the full posterior parameter distribution from the real data
full_param_df_hosp_only <- get_full_param_distrib(
  all_draws_hosp_only, static_params, vector_params, matrix_params
)

# Compare to the knows parameters
comp_df_hosp_only <- full_param_df_hosp_only %>%
  left_join(param_df,
    by = c("name", "index_rows", "index_cols")
  )

ggplot(comp_df_hosp_only %>% filter(name %in% c("rt", "p_hosp"))) +
  geom_line(aes(x = index_cols, y = value, group = draw), size = 0.05, alpha = 0.05) +
  geom_line(aes(x = index_cols, y = median), color = "black") +
  geom_line(aes(x = index_cols, y = true_value), color = "red") +
  facet_wrap(~name, scales = "free") +
  theme_bw() +
  xlab("Time (days)") +
  ylab("Parameter value") +
  scale_y_continuous(trans = "log") +
  ggtitle("Posterior parameter distribution vs known parameter values")
```
